---
title: "Introduction to the tidyverse: data manipulation with `dplyr`"
author: |
        | Christopher Skovron
        | Northwestern University 
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  revealjs::revealjs_presentation: 
    theme: simple
    highlight: haddock
    center: false
    transition: none
    #css: reveal.css
    self_contained: false
---

## dplyr: data manipulation 
      

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('~/Dropbox/tidyverse-workshop/')

```



## dplyr is the main workhorse of the tidyverse

This lesson draws on [Chapter 5 of R for Data Science](http://r4ds.had.co.nz/transform.html). 
For a condensed version, check out the [Introduction to dplyr](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html) vignette. 


## Example data: nycflights13

```{r}
#install.packages(c('nycflights13', 'tidyverse'))
library(nycflights13)
library(tidyverse)
```

## Note conflicts when you load the tidyverse

Take careful note of the conflicts message that's printed when you load the tidyverse. It tells you that dplyr overwrites some functions in base R. If you want to use the base version of these functions after loading dplyr, you'll need to use their full names: `stats::filter()` and `stats::lag()`.

## nycflights13

Example data: `nycflights13::flights`. This data frame contains all `r format(nrow(nycflights13::flights), big.mark = ",")` flights that departed from New York City in 2013. 

```{r}
flights
```


## It prints as a tibble 

You might also have noticed the row of three (or four) letter abbreviations under the column names. These describe the type of each variable:

* `int` stands for integers.

* `dbl` stands for doubles, or real numbers.

* `chr` stands for character vectors, or strings.

* `dttm` stands for date-times (a date + a time).

## Other kinds of variables you may see in a tibble

* `lgl` stands for logical, vectors that contain only `TRUE` or `FALSE`.

* `fctr` stands for factors, which R uses to represent categorical variables
  with fixed possible values.

* `date` stands for dates.

## dplyr verbs


* Pick observations by their values (`filter()`).
* Reorder the rows (`arrange()`).
* Pick variables by their names (`select()`).
* Create new variables with functions of existing variables (`mutate()`).
* Collapse many values down to a single summary (`summarise()`).


## We will frequently used these on grouped data 

These can all be used in conjunction with `group_by()` which changes the scope of each function from operating on the entire dataset to operating on it group-by-group. These six functions provide the verbs for a language of data manipulation.

## How dplyr verbs work 

All verbs work similarly: 

1.  The first argument is a data frame.

1.  The subsequent arguments describe what to do with the data frame,
    using the variable names (without quotes).
    
1.  The result is a new data frame.

Together these properties make it easy to chain together multiple simple steps to achieve a complex result. Let's dive in and see how these verbs work.

## Choose rows with `filter()`

For example, we can select all flights on January 1st with:

```{r}
filter(flights, month == 1, day == 1)
```

## Choose rows with `filter()`


When you run that line of code, dplyr executes the filtering operation and returns a new data frame. dplyr functions never modify their inputs, so if you want to save the result, you'll need to use the assignment operator, `<-`:

```{r}
jan1 <- filter(flights, month == 1, day == 1)
```



## Comparisons

To use filtering effectively, you have to know how to select the observations that you want using the comparison operators. R provides the standard suite: `>`, `>=`, `<`, `<=`, `!=` (not equal), and `==` (equal). 

Remember to use `=` instead of `==` when testing for equality. When this happens you'll get an informative error:

```{r, error = TRUE}
filter(flights, month = 1)
```


## Issues with floaitng point numbers 
There's another common problem you might encounter when using `==`: floating point numbers. These results might surprise you!

```{r}
sqrt(2) ^ 2 == 2
1/49 * 49 == 1
```


## Use `near()` instead
Computers use finite precision arithmetic (they obviously can't store an infinite number of digits!) so remember that every number you see is an approximation. Instead of relying on `==`, use `near()`:

```{r}
near(sqrt(2) ^ 2,  2)
near(1 / 49 * 49, 1)
```

## Logical operators

Multiple arguments to `filter()` are combined with "and": every expression must be true in order for a row to be included in the output. For other types of combinations, you'll need to use Boolean operators yourself: `&` is "and", `|` is "or", and `!` is "not". 

## Find all flights that departed in November or December:

```{r, eval = FALSE}
filter(flights, month == 11 | month == 12)
```


## Another way to write it using `%in%`

```{r, eval = FALSE}
nov_dec <- filter(flights, month %in% c(11, 12))
```

Sometimes you can simplify complicated subsetting by remembering De Morgan's law: `!(x & y)` is the same as `!x | !y`, and `!(x | y)` is the same as `!x & !y`. For example, if you wanted to find flights that weren't delayed (on arrival or departure) by more than two hours, you could use either of the following two filters:

```{r, eval = FALSE}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```



### Missing values

One important feature of R that can make comparison tricky are missing values, or `NA`s ("not availables"). `NA` represents an unknown value so missing values are "contagious": almost any operation involving an unknown value will also be unknown.

```{r}
NA > 5
10 == NA
NA + 10
NA / 2
```

The most confusing result is this one:

```{r}
NA == NA
```

It's easiest to understand why this is true with a bit more context:

```{r}
# Let x be Mary's age. We don't know how old she is.
x <- NA

# Let y be John's age. We don't know how old he is.
y <- NA

# Are John and Mary the same age?
x == y
# We don't know!
```

If you want to determine if a value is missing, use `is.na()`:

```{r}
is.na(x)
```

`filter()` only includes rows where the condition is `TRUE`; it excludes both `FALSE` and `NA` values. If you want to preserve missing values, ask for them explicitly:

```{r}
df <- tibble(x = c(1, NA, 3))
filter(df, x > 1)
filter(df, is.na(x) | x > 1)
```


## Arrange rows with `arrange()`

`arrange()` changes the order of rows order. It takes a data frame and a set of column names (or more complicated expressions) to order by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:

```{r}
arrange(flights, year, month, day)
```

## Arrange rows with `arrange()`


Use `desc()` to re-order by a column in descending order:

```{r}
arrange(flights, desc(arr_delay))
```


## Arrange rows with `arrange()`

Missing values are always sorted at the end:

```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
arrange(df, desc(x))
```

## Exercise
1.  Sort `flights` to find the fastest flights.

## Solution
```{r}
flights %>% 
  arrange(air_time)
```


## Select columns with `select()` 

`select()` chooses the columns you want out of the whole set of columns 

```{r}
# Select columns by name
select(flights, year, month, day)
```
## Select columns with `select()` 

```{r}
# Select all columns between year and day (inclusive)
select(flights, year:day)
```
## Select columns with `select()` 

```{r}
# Select all columns except those from year to day (inclusive)
select(flights, -(year:day))
```


## Helpers for `select()`

There are a number of helper functions you can use within `select()`:

* `starts_with("abc")`: matches names that begin with "abc".

* `ends_with("xyz")`: matches names that end with "xyz".

* `contains("ijk")`: matches names that contain "ijk".

* `matches("(.)\\1")`: selects variables that match a regular expression.
   This one matches any variables that contain repeated characters. You'll 
   learn more about regular expressions in [strings].
   
*  `num_range("x", 1:3)` matches `x1`, `x2` and `x3`.
   
   
## Use `rename()` to rename variables     

`select()` can be used to rename variables, but it's rarely useful because it drops all of the variables not explicitly mentioned. Instead, use `rename()`, which is a variant of `select()` that keeps all the variables that aren't explicitly mentioned:

## Use `rename()` to rename variables     

```{r}
rename(flights, tail_num = tailnum)
```

*Note that it's old_name  = new_name. This always throws me off.* 


## Add new variables with `mutate()`

Besides selecting sets of existing columns, it's often useful to add new columns that are functions of existing columns. That's the job of `mutate()`. 

## Make a smaller version of the dataset so we can see our work 

```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
```
## Add new variables with `mutate()`
```{r}
mutate(flights_sml,
  gain = arr_delay - dep_delay,
  speed = distance / air_time * 60
)
```
## Add new variables with `mutate()`

Note that you can refer to columns that you've just created:

```{r}
mutate(flights_sml,
  gain = arr_delay - dep_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```
## Add new variables with `mutate()`

If you only want to keep the new variables, use `transmute()`:

```{r}
transmute(flights,
  gain = arr_delay - dep_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```


There are many functions for creating new variables that you can use with `mutate()`. The key property is that the function must be vectorised: it must take a vector of values as input, return a vector with the same number of values as output. There's no way to list every possible function that you might use, but here's a selection of functions that are frequently useful:



*   Logical comparisons, `<`, `<=`, `>`, `>=`, `!=`, which you learned about
    earlier. If you're doing a complex sequence of logical operations it's 
    often a good idea to store the interim values in new variables so you can
    check that each step is working as expected.

## Useful helpers for `mutate()`

*   Ranking: there are a number of ranking functions, but you should 
    start with `min_rank()`. It does the most usual type of ranking 
    (e.g. 1st, 2nd, 2nd, 4th). The default gives smallest values the small
    ranks; use `desc(x)` to give the largest values the smallest ranks. 
    
    ```{r}
    y <- c(1, 2, 2, NA, 3, 4)
    min_rank(y)
    min_rank(desc(y))
    ```
## Useful helpers for `mutate()`
    
   If `min_rank()` doesn't do what you need, look at the variants
    `row_number()`, `dense_rank()`, `percent_rank()`, `cume_dist()`,
    `ntile()`.  See their help pages for more details.
    
    ```{r}
    row_number(y)
    dense_rank(y)
    percent_rank(y)
    cume_dist(y)
    ```



## Grouped summaries with `summarise()`

The last key verb is `summarise()`. It collapses a data frame to a single row:

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```


`summarise()` is not terribly useful unless we pair it with `group_by()`. This changes the unit of analysis from the complete dataset to individual groups. Then, when you use the dplyr verbs on a grouped data frame they'll be automatically applied "by group". For example, if we applied exactly the same code to a data frame grouped by date, we get the average delay per date:

```{r}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```




### Missing values

You may have wondered about the `na.rm` argument we used above. What happens if we don't set it?

```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

## Missing values

We get a lot of missing values! That's because aggregation functions obey the usual rule of missing values: if there's any missing value in the input, the output will be a missing value. Fortunately, all aggregation functions have an `na.rm` argument which removes the missing values prior to computation:

```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay, na.rm = TRUE))
```

## Missing values

In this case, where missing values represent cancelled flights, we could also tackle the problem by first removing the cancelled flights. We'll save this dataset so we can reuse in the next few examples.

```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

## Counts

Whenever you do any aggregation, it's always a good idea to include either a count (`n()`), or a count of non-missing values (`sum(!is.na(x))`). That way you can check that you're not drawing conclusions based on very small amounts of data. For example, let's look at the planes (identified by their tail number) that have the highest average delays:

```{r}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay)
  )

ggplot(data = delays, mapping = aes(x = delay)) + 
  geom_freqpoly(binwidth = 10)
```



## Exercise
1.  Find all flights that

    1. Had an arrival delay of two or more hours
    1. Flew to Houston (`IAH` or `HOU`)
    1. Were operated by United, American, or Delta
    1. Departed in summer (July, August, and September)
    
    


## Grouped summaries


##   Counts

  You've seen `n()`, which takes no arguments, and returns the 
    size of the current group. To count the number of non-missing values, use
    `sum(!is.na(x))`. To count the number of distinct (unique) values, use
    `n_distinct(x)`.
    
    ```{r}
    # Which destinations have the most carriers?
    not_cancelled %>% 
      group_by(dest) %>% 
      summarise(carriers = n_distinct(carrier)) %>% 
      arrange(desc(carriers))
    ```
    
##   Counts
    
Counts are so useful that dplyr provides a simple helper if all you want is 
    a count:
    
    ```{r}
    not_cancelled %>% 
      count(dest)
    ```

##   Counts
    
You can optionally provide a weight variable. For example, you could use 
    this to "count" (sum) the total number of miles a plane flew:
    
    ```{r}
    not_cancelled %>% 
      count(tailnum, wt = distance)
    ```
##   Counts
  
  Counts and proportions of logical values: `sum(x > 10)`, `mean(y == 0)`.
    When used with numeric functions, `TRUE` is converted to 1 and `FALSE` to 0. 
    This makes `sum()` and `mean()` very useful: `sum(x)` gives the number of 
    `TRUE`s in `x`, and `mean(x)` gives the proportion.
    
    ```{r}
    # How many flights left before 5am? (these usually indicate delayed
    # flights from the previous day)
    not_cancelled %>% 
      group_by(year, month, day) %>% 
      summarise(n_early = sum(dep_time < 500))
    
    # What proportion of flights are delayed by more than an hour?
    not_cancelled %>% 
      group_by(year, month, day) %>% 
      summarise(hour_perc = mean(arr_delay > 60))
    ```

## Grouping by multiple variables

When you group by multiple variables, each summary peels off one level of the grouping. That makes it easy to progressively roll up a dataset:

```{r}
daily <- group_by(flights, year, month, day)
(per_day   <- summarise(daily, flights = n()))
(per_month <- summarise(per_day, flights = sum(flights)))
(per_year  <- summarise(per_month, flights = sum(flights)))
```

Be careful when progressively rolling up summaries: it's OK for sums and counts, but you need to think about weighting means and variances, and it's not possible to do it exactly for rank-based statistics like the median. In other words, the sum of groupwise sums is the overall sum, but the median of groupwise medians is not the overall median.

## Ungrouping

If you need to remove grouping, and return to operations on ungrouped data, use `ungroup()`. 

```{r}
daily %>% 
  ungroup() %>%             # no longer grouped by date
  summarise(flights = n())  # all flights
```


## Grouped mutates (and filters)

Grouping is most useful in conjunction with `summarise()`, but you can also do convenient operations with `mutate()` and `filter()`:

##  Find the worst members of each group:

    ```{r}
    flights_sml %>% 
      group_by(year, month, day) %>%
      filter(rank(desc(arr_delay)) < 10)
    ```

##  Find all groups bigger than a threshold:

    ```{r}
    popular_dests <- flights %>% 
      group_by(dest) %>% 
      filter(n() > 365)
    popular_dests
    ```

##  Standardise to compute per group metrics:

    ```{r}
    popular_dests %>% 
      filter(arr_delay > 0) %>% 
      mutate(prop_delay = arr_delay / sum(arr_delay)) %>% 
      select(year:day, dest, arr_delay, prop_delay)
    ```

## Exercise

1.  Come up with another approach that will give you the same output as 
    `not_cancelled %>% count(dest)` and 
    `not_cancelled %>% count(tailnum, wt = distance)` (without using 
    `count()`).

1.  Which carrier has the worst delays? Challenge: can you disentangle the
    effects of bad airports vs. bad carriers? Why/why not? (Hint: think about
    `flights %>% group_by(carrier, dest) %>% summarise(n())`)

## Advanced verbs: `summarize_all()`, `summarize_at()`, `summarize_if()`

The `_all`, `_at` and `_if` suffixes work on `mutate`, `summarize`, and `transmute`. (British English speakers, you can use `summarise` too.)

- `_all` applies the operation to all columns
- `_at` applies the operation only to named columns
- `_if` applies the operation if some condition is true 


## `summarize_all()`
    ```{r}
iris %>%
  group_by(Species) %>%
  summarise_all(mean)
```

## `summarize_at()`

```{r}
starwars %>% 
summarise_at(c("height", "mass"), mean, na.rm = TRUE)

```

## `mutate_at()` with a helper to choose columns

```{r}
iris %>% 
  mutate_at(vars(matches("Sepal")), log)
```

## `summarise_if`

```{r}
# The _if() variants apply a predicate function (a function that
# returns TRUE or FALSE) to determine the relevant subset of
# columns. Here we apply mean() to the numeric columns:
starwars %>% summarise_if(is.numeric, mean, na.rm = TRUE)
#> # A tibble: 1 x 3
#>   height  mass birth_year
#>    <dbl> <dbl>      <dbl>
#> 1   174.  97.3       87.6
```












