---
title: "Exercises Part 2: wrangle with readr and tidyr"
author: ""
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---

```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
library(tidyverse)
```


## Exercises: tibbles 

1.  How can you tell if an object is a tibble? (Hint: try printing `mtcars`,
    which is a regular data frame). 

1.  Compare and contrast the following operations on a `data.frame` and 
    equivalent tibble. What is different? Why might the default data frame
    behaviours cause you frustration?
    
    ```{r, eval = FALSE}
    df <- data.frame(abc = 1, xyz = "a")
    df$x
    df[, "xyz"]
    df[, c("abc", "xyz")]
    ```

1.  If you have the name of a variable stored in an object, e.g. `var <- "mpg"`,
    how can you extract the reference variable from a tibble?

1.  Practice referring to non-syntactic names in the following data frame by:

    1.  Extracting the variable called `1`.

    1.  Plotting a scatterplot of `1` vs `2`.

    1.  Creating a new column called `3` which is `2` divided by `1`.
        
    1.  Renaming the columns to `one`, `two` and `three`. 
    
    ```{r}
    annoying <- tibble(
      `1` = 1:10,
      `2` = `1` * 2 + rnorm(length(`1`))
    )
    ```

1.  What does `tibble::enframe()` do? When might you use it?

1.  What option controls how many additional column names are printed
    at the footer of a tibble?

    
### Exercises: readr

1.  What function would you use to read a file where fields were separated with  
    "|"?
    
1.  Apart from `file`, `skip`, and `comment`, what other arguments do
    `read_csv()` and `read_tsv()` have in common?
    
1.  What are the most important arguments to `read_fwf()`?
   
1.  Sometimes strings in a CSV file contain commas. To prevent them from
    causing problems they need to be surrounded by a quoting character, like
    `"` or `'`. By convention, `read_csv()` assumes that the quoting
    character will be `"`, and if you want to change it you'll need to
    use `read_delim()` instead. What arguments do you need to specify
    to read the following text into a data frame?
    
    ```{r, eval = FALSE}
    "x,y\n1,'a,b'"
    ```
    
1.  Identify what is wrong with each of the following inline CSV files. 
    What happens when you run the code?
    
    ```{r, eval = FALSE}
    read_csv("a,b\n1,2,3\n4,5,6")
    read_csv("a,b,c\n1,2\n1,2,3,4")
    read_csv("a,b\n\"1")
    read_csv("a,b\n1,2\na,b")
    read_csv("a;b\n1;3")
    ```
    
    
    
    
    