---
title: "Exercises Part 2: wrangle with readr and tidyr"
author: ""
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---

```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
library(tidyverse)
```


## Exercises: tibbles 

1.  How can you tell if an object is a tibble? (Hint: try printing `mtcars`,
    which is a regular data frame). 

1.  Compare and contrast the following operations on a `data.frame` and 
    equivalent tibble. What is different? Why might the default data frame
    behaviours cause you frustration?
    
    ```{r, eval = FALSE}
    df <- data.frame(abc = 1, xyz = "a")
    df$x
    df[, "xyz"]
    df[, c("abc", "xyz")]
    ```

1.  If you have the name of a variable stored in an object, e.g. `var <- "mpg"`,
    how can you extract the reference variable from a tibble?

1.  Practice referring to non-syntactic names in the following data frame by:

    1.  Extracting the variable called `1`.

    1.  Plotting a scatterplot of `1` vs `2`.

    1.  Creating a new column called `3` which is `2` divided by `1`.
        
    1.  Renaming the columns to `one`, `two` and `three`. 
    
    ```{r}
    annoying <- tibble(
      `1` = 1:10,
      `2` = `1` * 2 + rnorm(length(`1`))
    )
    ```

1.  What does `tibble::enframe()` do? When might you use it?

1.  What option controls how many additional column names are printed
    at the footer of a tibble?

    
### Exercises: readr

1.  What function would you use to read a file where fields were separated with  
    "|"?
    
1.  Apart from `file`, `skip`, and `comment`, what other arguments do
    `read_csv()` and `read_tsv()` have in common?
    
1.  What are the most important arguments to `read_fwf()`?
   
1.  Sometimes strings in a CSV file contain commas. To prevent them from
    causing problems they need to be surrounded by a quoting character, like
    `"` or `'`. By convention, `read_csv()` assumes that the quoting
    character will be `"`, and if you want to change it you'll need to
    use `read_delim()` instead. What arguments do you need to specify
    to read the following text into a data frame?
    
    ```{r, eval = FALSE}
    "x,y\n1,'a,b'"
    ```
    
1.  Identify what is wrong with each of the following inline CSV files. 
    What happens when you run the code?
    
    ```{r, eval = FALSE}
    read_csv("a,b\n1,2,3\n4,5,6")
    read_csv("a,b,c\n1,2\n1,2,3,4")
    read_csv("a,b\n\"1")
    read_csv("a,b\n1,2\na,b")
    read_csv("a;b\n1;3")
    ```
    
    
    
### Exercises: tidyr

1.  Using prose, describe how the variables and observations are organised in
    each of the sample tables.

1.  Compute the `rate` for `table2`, and `table4a` + `table4b`. 
    You will need to perform four operations:

    1.  Extract the number of TB cases per country per year.
    1.  Extract the matching population per country per year.
    1.  Divide cases by population, and multiply by 10000.
    1.  Store back in the appropriate place.
    
    Which representation is easiest to work with? Which is hardest? Why?

1.  Recreate the plot showing change in cases over time using `table2`
    instead of `table1`. What do you need to do first?
    
    
### Exercises: tidyr

1.  Why are `gather()` and `spread()` not perfectly symmetrical?  
    Carefully consider the following example:
    
    ```{r, eval = FALSE}
    stocks <- tibble(
      year   = c(2015, 2015, 2016, 2016),
      half  = c(   1,    2,     1,    2),
      return = c(1.88, 0.59, 0.92, 0.17)
    )
    stocks %>% 
      spread(year, return) %>% 
      gather("year", "return", `2015`:`2016`)
    ```
    
    (Hint: look at the variable types and think about column _names_.)
    
    Both `spread()` and `gather()` have a `convert` argument. What does it 
    do?

1.  Why does this code fail?

    ```{r, error = TRUE}
    table4a %>% 
      gather(1999, 2000, key = "year", value = "cases")
    ```

1.  Why does spreading this tibble fail? How could you add a new column to fix
    the problem?

    ```{r}
    people <- tribble(
      ~name,             ~key,    ~value,
      #-----------------|--------|------
      "Phillip Woods",   "age",       45,
      "Phillip Woods",   "height",   186,
      "Phillip Woods",   "age",       50,
      "Jessica Cordero", "age",       37,
      "Jessica Cordero", "height",   156
    )
    ```

1.  Tidy the simple tibble below. Do you need to spread or gather it?
    What are the variables?

    ```{r}
    preg <- tribble(
      ~pregnant, ~male, ~female,
      "yes",     NA,    10,
      "no",      20,    12
    )
    ```
    
### Exercises

1.  What do the `extra` and `fill` arguments do in `separate()`? 
    Experiment with the various options for the following two toy datasets.
    
    ```{r, eval = FALSE}
    tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
      separate(x, c("one", "two", "three"))
    
    tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
      separate(x, c("one", "two", "three"))
    ```

1.  Both `unite()` and `separate()` have a `remove` argument. What does it
    do? Why would you set it to `FALSE`?

1.  Compare and contrast `separate()` and `extract()`.  Why are there
    three variations of separation (by position, by separator, and with
    groups), but only one unite?
    