---
title: "Exercises Part 3: stringr"
author: ""
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---

```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
library(tidyverse)
```

### Exercises

1.  In code that doesn't use stringr, you'll often see `paste()` and `paste0()`.
    What's the difference between the two functions? What stringr function are
    they equivalent to? How do the functions differ in their handling of 
    `NA`?
    
1.  In your own words, describe the difference between the `sep` and `collapse`
    arguments to `str_c()`.

1.  Use `str_length()` and `str_sub()` to extract the middle character from 
    a string. What will you do if the string has an even number of characters?

1.  What does `str_wrap()` do? When might you want to use it?

1.  What does `str_trim()` do? What's the opposite of `str_trim()`?

### Exercises

1. Find all words that come after a "number" like "one", "two", "three" etc.
   Pull out both the number and the word.

1. Find all contractions. Separate out the pieces before and after the 
   apostrophe.

### Exercises

1.   Replace all forward slashes in a string with backslashes.

1.   Implement a simple version of `str_to_lower()` using `replace_all()`.

1.   Switch the first and last letters in `words`. Which of those strings
     are still words?

#### Exercises

1.  Split up a string like `"apples, pears, and bananas"` into individual
    components.
    
1.  Why is it better to split up by `boundary("word")` than `" "`?

1.  What does splitting with an empty string (`""`) do? Experiment, and
    then read the documentation.
    
### Exercises

1.  How would you find all strings containing `\` with `regex()` vs.
    with `fixed()`?

1.  What are the five most common words in `sentences`?