---
title: "Exercises Part 4: Dates with lubridate"
author: ""
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---

```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
library(tidyverse)
```

### Exercises

1.  What happens if you parse a string that contains invalid dates?

    ```{r, eval = FALSE}
    ymd(c("2010-10-10", "bananas"))
    ```

1.  What does the `tzone` argument to `today()` do? Why is it important?

1.  Use the appropriate lubridate function to parse each of the following dates:

    ```{r}
    d1 <- "January 1, 2010"
    d2 <- "2015-Mar-07"
    d3 <- "06-Jun-2017"
    d4 <- c("August 19 (2015)", "July 1 (2015)")
    d5 <- "12/30/14" # Dec 30, 2014
    ```
    
    

### Exercises

1.  How does the distribution of flight times within a day change over the 
    course of the year?
    
1.  Compare `dep_time`, `sched_dep_time` and `dep_delay`. Are they consistent?
    Explain your findings.

1.  Compare `air_time` with the duration between the departure and arrival.
    Explain your findings. (Hint: consider the location of the airport.)
    
1.  How does the average delay time change over the course of a day?
    Should you use `dep_time` or `sched_dep_time`? Why?

1.  On what day of the week should you leave if you want to minimise the
    chance of a delay?

1.  What makes the distribution of `diamonds$carat` and 
    `flights$sched_dep_time` similar?

1.  Confirm my hypothesis that the early departures of flights in minutes
    20-30 and 50-60 are caused by scheduled flights that leave early. 
    Hint: create a binary variable that tells you whether or not a flight 
    was delayed.
    

### Exercises

1.  Why is there `months()` but no `dmonths()`?

1.  Explain `days(overnight * 1)` to someone who has just started 
    learning R. How does it work?

1.  Create a vector of dates giving the first day of every month in 2015.
    Create a vector of dates giving the first day of every month
    in the _current_ year.

1.  Write a function that given your birthday (as a date), returns 
    how old you are in years.

1.  Why can't `(today() %--% (today() + years(1)) / months(1)` work?
 of all time zone names with `OlsonNames()`: